Prediction Markets for Exponential Families

### 2014-01-31 WriteUp/Original/ExpFly.tex
tidied up main folder to just have submission doc

% Adjusted to add \usepackage[numbers]{natbib}
% \bibliographystyle{acmsmall} and \documentclass[prodmode,acmec]{ec-acmsmall}
% Jan 5, 2013 - David Parkes
%
% v2-acmsmall-sample.tex, dated March 6 2012
% This is a sample file for ACM small trim journals
%
% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.3 - March 2012

\documentclass[prodmode,acmec]{ec-acmsmall} % Aptara syntax

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2013}
\acmMonth{6}
\usepackage[numbers]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{amsthm}
\usepackage{parskip}
\usepackage{booktabs}

%new commands
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\mX}{\mathcal{X}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\grad}{\nabla}
\newcommand{\Exp}{\mathbf{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\mM}{\ensuremath{\mathcal{M}}}
\newcommand{\hmu}{\ensuremath{\hat{\mu}}}
\newcommand{\mP}{\ensuremath{\mathcal{P}}}
\newcommand{\mY}{\ensuremath{\mathcal{Y}}}
\newcommand{\mS}{\ensuremath{\mathcal{S}}}
\newcommand{\ds}{\displaystyle}
\newcommand{\mF}{\mathcal{F}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\htheta}{\hat{\theta}}
%---------------------------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{con}[thm]{Conjecture}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{ex}[thm]{Example}
\newtheorem{qn}[thm]{Question}
\newcommand{\defn}{\bigbreak\noindent{\bf Definition. }}
\newcommand{\undefn}{\bigbreak}
\newcommand{\rmk}{\bigbreak\noindent{\bf Remark. }}
\newcommand{\unrmk}{\bigbreak}
%\newcommand{\proof}{\noindent{\bf Proof. }}
\newcommand{\unproof}{\hfill$\Box$\bigbreak}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\holds}[1]{\stackrel{?}{#1}}
\newcommand{\conv}{\mathrm{conv}}
\newcommand{\ie} {{\it i.e. }}
\newcommand{\eg} {{\it e.g. }}
\newcommand{\etal} {{\it et al. }}
\newcommand{\E}{\mathbf{E}}
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\family}{\mathcal F} 
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\muvec}{\pmb{\mu}}

% Document starts

\begin{document}
% Page heads
\markboth{Abernethy et al.}{Maximum Entropy Prediction Markets}
% Title portion
\title{Maximum Entropy Prediction Markets}
\author{JACOB ABERNETHY
\affil{University of Michigan, Ann Arbor}
SINDHU KUTTY
\affil{University of Michigan, Ann Arbor}
S\'{E}BASTIEN LAHAIE
\affil{Microsoft Research, New York City}
RAHUL SAMI
\affil{University of Michigan, Ann Arbor}}
%%\author{GANG ZHOU
%%\affil{College of William and Mary}
%%YAFENG WU
%%\affil{University of Virginia}
%%TING YAN
%%\affil{Eaton Innovation Center}
%%TIAN HE
%%\affil{University of Minnesota}
%%CHENGDU HUANG
%%\affil{Google}
%%JOHN A. STANKOVIC
%%\affil{University of Virginia}
%%TAREK F. ABDELZAHER
%%\affil{University of Illinois at Urbana-Champaign}}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

\begin{abstract}
In this paper, we draw connections between the aggregation performed by learning algorithms and the information aggregation done in prediction markets. We show that, under reasonable conditions, the behavior of rational traders can be understood as the result of performing a learning algorithm on their private data. Similarly,  the market state can be interpreted as a distribution over the outcome space. In particular, we show that a proper scoring rule can be derived from maximum entropy distributions. This scoring rule can be used as a general form of LMSR in prediction markets with over continuous outcome spaces. In order to provide insight on the behavior of rational traders in the market, we use the concept of exponential utility. We show that the traders' behavior can be understood as updating his belief using a Bayesian process and updating the market state in accordance with this utility function. These maxent prediction  markets can also be used to design  markets that are robust against adversarial traders. In fact, when traders are required to report their budgets and their beliefs, we can show that an informative trader eventually makes money and damaging traders eventually have limited influence in the market. Using ideas from convex analysis and the properties of the prediction market, we analyze the properties of the maxent market maker thus providing insight into the information content of the prediction market.
\end{abstract}

\category{J.4}{Social and Behavioral Sciences}{Economics}
\category{I.2.6} {Artificial Intelligence}{Learning}

\terms{Prediction Market Design, Machine Learning Algorithms}

\keywords{Exponential Families, Bayesian Learning, Maximum Entropy Distributions, generalized LMSR}

\acmformat{Jacob Abernethy, Sindhu Kutty, S\'{e}bastien Lahaie, and Rahul Sami, 2014. Maximum Entropy Prediction Markets}
%%Gang Zhou, Yafeng Wu, Ting Yan, Tian He, Chengdu Huang, John A. Stankovic, and Tarek F. Abdelzaher, 2010. A multifrequency MAC specially designed for  wireless sensor network applications.

% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

\begin{bottomstuff}
%%This work is supported by the National Science Foundation, under
%%grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.
%%
%%Author's addresses: G. Zhou, Computer Science Department,
%%College of William and Mary; Y. Wu  {and} J. A. Stankovic,
%%Computer Science Department, University of Virginia; T. Yan,
%%Eaton Innovation Center; T. He, Computer Science Department,
%%University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
%%(Current address) NASA Ames Research Center, Moffett Field, California 94035.
\end{bottomstuff}

\maketitle

\section{Introduction}
%------------------------------------------------------------
%Outline
%------------------------------------------------------------
%- exp family review
%- maxent scoring rule
%	- generalized lmsr
%	- does modularity (a la Hanson) still hold?
%- prediction market setup
%	- MLE with risk neutrality
%- conjugate prior market
%	- also on hyperparam?
%- exp-util
%- equilibrium for repeated trades
%- budgets and adversarial markets

We have dual goals in this paper. On the one hand, we highlight a structural similarity between prediction markets and exponential families. We see this as the syntax of our prediction market mechanism. On the other, this formulation has rich semantics as well: it allows for analysis of market behavior under various environments. For instance, we analyze market behavior with budget limited traders, traders that are exponential-utility-maximizers.
\section{Notation and Definitions}
convex conjugates, bregman divergences, scoring rules, sufficient statitics, exp families, Bayesian updates, conjugate prior.
\section{Maximum Likelihood Scoring Rules}

%%%
%%%
%%%

\subsection{Preliminaries}

We consider a state space $\mX$ which may be discrete or continuous, and a measure $\nu$ over sets of states. We will typically take the state space to be  $\bR$ or $\bR^d$ together with the Lebesgue measure (perhaps restricted to some subset of states), or some finite set together with the counting measure.\footnote{We will not need to make reference to the underlying $\sigma$-algebra, but to complete the specification of the measure space we take it to be the collection of Borel sets in the case of $\bR$ and $\bR^d$, and the power set in the case of a discrete set.} A \emph{probability density} is a $\nu$-measurable function $p : \mX \rightarrow \bR_+$ that integrates to 1 over $\mX$.

Let $\mP$ denote the set of all probability densities over the given measure space. We are interested in eliciting information about some unknown density $p \in \mP$ which corresponds to the beliefs of an expert. The expert may be familiar with $p$ in its entirety, or only with certain properties of $p$.  Given a set of densities $\mD \subseteq \mP$, a \emph{property}  is a real (possibly vector-valued) function $\Gamma : \mD \rightarrow \bR^k$. With a slight abuse of terminology we call elements of its image $\Theta = \Gamma(\mD)$ \emph{properties} and we say that $p$ has property $\theta$ when $\Gamma(p) = \theta$. If $\Gamma$ is one-to-one, we instead call $\Gamma$ a \emph{parametrization} and refer to the elements of its image $\Theta$ as \emph{parameters}.

To be concrete, our outcome space could for instance be $\bR$ with $\nu$ the Lebesgue measure restricted to $[0,+\infty)$. The set $\mP$ then consists of all probability densities with support contained in the non-negative reals, and $\mD \subset \mP$ might consist of the densities of exponential distributions. An example of a property over $\mP$ is the mapping from a density to its mean.\footnote{The property may not be defined for all $p \in \mP$. In this case we implicitly restrict it to the subset of $\mP$ for which it is defined whenever its domain $\mD$ is not made explicit.} When restricted to the exponential densities $\mD$, this mapping is a parametrization. In much of this work, as in this example, the purpose of the measure $\nu$ is to implicitly encode bounds on the support of possible densities. The reason we use this approach, rather than explicitly stating bounds, is that it allows for clearer connections with maximum entropy estimation later on.

Given the set of relevant densities $\mD \subseteq \mP$, which contains the expert's belief $p$, and a property $\Gamma$ over this set, the expert is asked to report $\theta = \Gamma(p)$. The expert is rewarded according to a \emph{scoring rule} $S : \Theta \times \mX \rightarrow \bR \cup \{-\infty\}$ which pays it $S(\htheta,x)$ according to how well its report $\htheta \in \Theta$ agrees with the eventual outcome $x \in \mX$. The idea is to design the scoring rule so that the expert is incentivized to report its true belief, setting $\htheta = \theta$. 
%
\begin{definition}
Given a property $\Gamma : \mD \rightarrow \bR^k$, a scoring rule $S$ is  \emph{proper} over domain $\mD$ if for each $\theta \in \Theta$ and $p \in \mD$ with property $\theta$, we have
%
\begin{equation} \label{eq-proper}
\Exp_p[S(\theta,x)] > \Exp_p[S(\htheta,x)]
\end{equation}
%
for all $\htheta \neq \theta$. If the domain is $\mD = \mP$, the set of all possible densities, then we say the scoring rule is \emph{strongly proper}.
\end{definition}
%

\begin{itemize}
\item Note that the inequality is strict, in the literature these rules are called \emph{strictly} proper, but we only consider strictly proper rules here so don't make this qualification.
\item Lambert et al.'s rules are actually strongly proper, however we consider scoring rules over vector valued properties. Lambert et al only develop these by adding together different scoring rules for each property. We go further in our constructions and develop rules that are not necessarily separable.
\item It is useful to consider rules that are not strongly proper because we might have constructions for these for properties where no strongly proper rule is known.
\item It is important to understand the informational requirements placed on the expert. Under a proper scoring rule, the expert does not need to be aware of the entire density $p$ to be incentivized to report its property $\theta$, since $\theta$ is an optimal report no matter what the density might be. The agent only needs to agree that the density indeed lies in $\mD$.
\item For instance, if we are considering densities with support on $[0,+\infty)$, and $\mD$ is the set of lognormal densities, then a proper scoring rule over $\mD$ for the mean does not require the agent to know anything beyond the mean (e.g., it does not need to know the variance).  
\item A strongly proper scoring rule goes further; with such a rule we do not require the expert to believe the density is drawn from some subset $\mD$, only that it agrees with the support implied by the base measure $\nu$. Therefore, a strongly proper scoring rule for the mean here would elicit the mean no matter what kind of density over the positive reals might apply (e.g., exponential, Pareto, lognormal).
\item In practical scenarios this seems like an important extension, because while it might be clear which states are possible, it might be difficult to ensure the expert agrees that the probability density for states falls within some restricted family $\mD$.
\end{itemize}

%%%
%%%
%%%

\subsection{Maximum Likelihood}

%
\begin{theorem} \label{maxlike}
The logarithmic scoring rule defined by
%
\[ S(\theta, x) = \log p(x;\theta) \]
%
is weakly proper for $\mF$ over $\Theta$ if and only if the maximum likelihood estimate for $\theta$ is consistent. Furthermore, if $\mM$ is an alternative parameter space such that there exists a bijection $t : \mM \rightarrow \Theta$, then $\bar{S}(\mu, x) = S(t(\mu), x)$ is weakly proper for $\mF$ over $\mM$. 
\end{theorem}
%

%
\begin{example}
Suppose $X$ is supported on $[0,+\infty)$ and follows an exponential distribution with density $f(x;\lambda) = \lambda e^{-\lambda x}$ parametrized by a rate $\lambda > 0$. The mean $\mu$ is related to the rate via the one-to-one mapping $\mu = 1/\lambda$, so the density can alternatively be parametrized by the mean. According to Theorem~\ref{maxlike} the following is a weakly proper scoring rule for the mean of an exponential distribution:
%
\begin{equation} \label{exp-scoring}
S(\mu, x) = -\frac{x}{\mu} - \log \mu.
\end{equation}
%
To verify this, suppose the agent believes $X$ follows an exponential distribution with rate $\lambda_0$, equivalently mean $\mu_0 = 1/\lambda_0$. By the one-to-one relation between rate and mean, choosing $\mu$ to maximize the expected score~(\ref{exp-scoring}) is equivalent to choosing $\lambda$ to maximize $\Exp{-\lambda x + \log \lambda} = -\lambda \mu_0 + \log \lambda$. The latter is strictly concave in $\lambda$, and using basic calculus its unique global maximum is $\lambda^* = 1/\mu_0$, which therefore leads the agent to report $\mu = \mu_0 = 1/\lambda^*$ under scoring rule~(\ref{exp-scoring}). 
\end{example}
%

%
\begin{example}
Suppose that $X$ is supported on $[1,+\infty)$ and follows a Pareto distribution with density $f(x;\alpha) = \alpha/x^{\alpha+1}$ parametrized by an index $\alpha > 0$. The mean $\mu$ is related to the index via the one-to-one mapping $\mu = \frac{\alpha}{\alpha-1}$, so the density can alternatively be parametrized by the mean. Theorem~\ref{maxlike} gives the following weakly proper scoring rule for the mean of a Pareto distribution with support on $[1,+\infty)$:
%
\begin{equation} \label{pareto-scoring}
S(\mu,x) = \log \frac{\mu}{\mu-1} - \left(\frac{\mu}{\mu-1} + 1\right) \log x.
\end{equation}
%
To verify this, suppose the agent believes $X$ follows a Pareto distribution with index $\alpha_0$, equivalently mean $\mu_0 = \frac{\alpha_0}{\alpha_0-1}$. By the one-to-one relation between the index and mean, choosing $\mu$ to maximize the expected score~(\ref{pareto-scoring}) is equivalent to choosing $\alpha$ to maximize $\Exp{\log \alpha - (\alpha+1)\log x} = \log \alpha - (\alpha+1)\Exp{\log x}$. The latter is strictly concave in $\alpha$, and using basic calculus its unique global maximum is $\alpha^* = 1/\Exp{\log X} = \alpha_0$, where the last equality follows from the fact that $\log X$ has an exponential distribution with rate $\alpha$ when $X$ is Pareto with rate $\alpha$. Therefore~(\ref{pareto-scoring}) leads the agent to report $\mu = \mu_0 = \frac{\alpha^*}{\alpha^* - 1}$. 
\end{example}
%

>>>>

\documentclass{article}
%\usepackage{mystyle}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{graphics}
\usepackage{enumerate}
\usepackage{amstext}

\newcommand{\xvec}{\mathbf{x}} %outcome values %skk changed to xvec from betavec
\newcommand{\family}{\mathcal F} 
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\muvec}{\pmb{\mu}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\eps}{\mbox{$\epsilon$}}
\newcommand{\phantomset}{\mathbb{H}}
\newcommand{\pexp}{\overline{e}}
\newcommand{\aexp}{\tilde{e}}

\newcommand{\hyperfamily}{\mathcal{F}^{*}}
\newcommand{\hyperP}{P^{*}}
\newcommand{\hyperb}{\mathbf{b}}
\newcommand{\hyperm}{\mathbf{m}}
\newcommand{\hyperbeta}{\pmb{\beta}^{*}}
\newcommand{\hyperlpf}{\mathbf{\psi}^{*}}
\newcommand{\zvec}{\mathbf{z}} %% \hyperbeta is decomposed as (r\zvec,r)
\newcommand{\pvec}{\mathbf{p}}
%XXX agents and data
\newcommand{\honestset}{\mathcal{H}}
\newcommand{\attackset}{\overline{\mathcal{H}}}
\newcommand{\xseq}{\mathbf{x}}
\newcommand{\att}{\mathcal{A}}
\newcommand{\nullatt}{\mathcal{A}_\phi}
\newcommand{\xt}{\tilde{\mathbf{x}}}
%notation: \xt_\att(\xseq) is attacked seq
%\newcommand{\bt}{\tilde{\mathbf{b}}} %attack vector in conj prior section
\newcommand{\Pt}{\tilde{P}} %attack hyperdistribution in conj prior section
%XXX algorithms and predictions
\newcommand{\pred}{Q}
%learning algorithm: Z
%omniscient/optimal learning algorithm: Z_O
%XXX losses and gains, informativeness, regret
% L(\pred_k, \xvec_k) : loss from pred when true outcome is x
% L(\pred_k, \betavec_k) : loss in revealed parameter model
% G_Z(u_i, \xt, \xvec_k) :
% G_Z (u_i, \xt): expected loss, under prior conditioned by \xseq
\newcommand{\info}{h} %% use \info_{ik}
\newcommand{\reg}{\mbox{Reg}} % use \reg(Z)
\newcommand{\ureg}{\mbox{UR}} % unaccounted regret, used later;%XXX  WTM/ILS notation
\newcommand{\Inf}[2]{y_{{#1}{#2}}} %{Y_{#1}^{(#2)}}
\newcommand{\Infvec}{\mathbf{y}}
\newcommand{\Gain}[2]{G_{{#1}{#2}}} %{G_{#1}^{(#2)}}
\newcommand{\Rep}[2]{r_{{#1}{#2}}}
\newcommand{\Iik}{\Inf{i}{k}}
\newcommand{\Gik}{\Gain{i}{k}}
\newcommand{\Ghik}{\hat{G}_{ik}} %estimate of unnormalized gain
\newcommand{\rik}{\Rep{i}{k}}
%\newcommand{\xik}{x_{ik}}
\newcommand{\yb}{\overline{y}}
%\newcommand{\ubar}{{\overline{u}}}

% G(\yvec, u_i, \xt, \xvec_k)
% G(y_i, u_i, \xt, \xvec_k)
% B(r): budget associated with reputation r

%XXX assorted notation
\newcommand{\var}{\mbox{Var}}
\newcommand{\IL}{\mbox{IL}_i} %% in its use, always refers to user i
\newcommand{\rinit}{\Rep{i}{1}} % in its use in IL proof, always refers to user i

\newcommand{\ddG}[1]{\frac{d#1}{dG}}
\newcommand{\sdG}[1]{\frac{d^2 #1}{dG^2}}
\newcommand{\ddg}[1]{\frac{d#1}{dg}}
\newcommand{\sdg}[1]{\frac{d^2 #1}{dg^2}}
\newcommand{\ddu}{\frac{d}{du}}
\newcommand{\sdu}{\frac{d^2}{du^2}}

\newcommand{\hk}{\info_{ik}} %shorthand used in info loss section
\newcommand{\htt}{h_t}
\newcommand{\Hk}{H_{ik}}
\newcommand{\Ht}{H_t}
\newcommand{\Gk}{G_k}
\newcommand{\gk}{g_k}
\newcommand{\gt}{g_t}
\newcommand{\kmin}{\overline{k}}
\newcommand{\Hmin}{\overline{H}}
\newcommand{\HM}{H_M}

\newcommand{\GFe}{\mbox{$\mathrm{GF}_\epsilon$}}
\newcommand{\EGFe}{\mbox{$\mathrm{EGF}_\epsilon$}}
\newcommand{\pihat}{\hat{\pi}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\uvec}{\mathbf{u}}
\newcommand{\wvec}{\mathbf{w}}
\newcommand{\qlim}{\tilde{q}}
\newcommand{\qbar}{\overline{q}}
\newcommand{\Gt}{\~{G}}
\newcommand{\seq}{\sigma}
\newcommand{\EG}{\mbox{EG}}
\newcommand{\avec}{\mathbf{a}}
\newcommand{\xbar}{{\overline{x}}}
\newcommand{\yhat}{\hat{y}}
\newcommand{\what}{\hat{w}}
\newcommand{\gik}{g_{ik}}
\newcommand{\xtilde}{\tilde{x}}

%margin comments in red
\newcommand{\com}[1]{\marginpar{\textcolor{red}{#1}}}

%\newcommand{\xvec}{\mathbf{x}}
\newcommand{\yvec}{\mathbf{y}}
\newcommand{\Gauss}{\mathcal{N}}
\newcommand{\Xvec}{\mathbf{X}}
%\newcommand{\qvec}{\mathbf{q}}
\newcommand{\wvechat}{\hat{\mathbf{w}}}
\newcommand{\Gbar}{\overline{G}}
\newcommand{\Lmat}{\mbox{$\mathbf{L}$}}
\newcommand{\Hmat}{\mbox{$\mathbf{H}$}}
\newcommand{\Kmat}{\mbox{$\mathbf{K}$}}
\newcommand{\Amat}{\mbox{$\mathbf{A}$}}
\newcommand{\etal}{{\it et al.}}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand\beqn{\begin{eqnarray*}}
\newcommand\eeqn{\end{eqnarray*}}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\eqncom}[1]{\quad{\mbox{\texttt{(#1)}}}}

\newenvironment{proof*}{\noindent{\bf Proof:}}{}
\newcommand{\note}[1]{\textcolor{gray}{#1}}
\newcommand{\ccite}[1]{\textcolor{blue}{#1}}

 \newtheorem{theorem}{Theorem}
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{prop}[theorem]{Proposition}
 \newtheorem{definition}[theorem]{Definition}
 \newtheorem{claim}[theorem]{Claim}
 \newtheorem{corollary}[theorem]{Corollary}
 \newcommand{\qed}{\hfill\rule{7pt}{7pt}}
\newenvironment{proof}{\noindent{\bf Proof:}}{\qed\medskip}

% If your paper is accepted, change the options for the package
% aistats2e as follows:
%
%\usepackage[accepted]{aistats2e}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

\begin{document}
% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
\title{Prediction Markets for Exponential Families}
\author{ Sindhu Kutty\\ Department of Computer Science and Engineering\\University of Michigan, Ann Arbor
\and
Rahul Sami \\School of Information\\University of Michigan, Ann Arbor} 
\date\today
\maketitle
\begin{abstract}
  %The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
%  both left and right-hand margins. Use 10~point type, with a vertical
%  spacing of 11~points.  The {\bf Abstract} heading must be centered,
%  bold, and in point size 12. Two line spaces precede the
%  Abstract. The Abstract must be limited to one paragraph.
 This paper identifies and develops new connections between two approaches to aggregating information from multiple sources: statistical learning techniques and prediction markets that rely on profit-seeking traders to update prices.
 We present a new rich class of prediction markets for infinite outcome spaces that are based on parametric distribution families. For these markets, we prove that a sequence of traders maximizing their expected profit will result in price sequences analogous to sequential maximum likelihood estimation from statistics. Further, we show that even when some traders are adversarial attackers rather strategic profit-maximizers, these prediction markets continue to have attractive learning properties: The damage that the attackers can cause is bounded by their initial budgets, and unlike attackers causing damage, informative traders' budgets will grow in expectation as they trade in more markets.
\end{abstract}

\section{INTRODUCTION}
Prediction markets are aggregation mechanisms that allow market prices to be interpreted as predictive probabilities on an event. Each trader in the market is assumed to have some private information that he uses to make a prediction on the outcome of the event. The traders are assumed to be strategic in the sense that every trader wants to maximize his profit by participating in the market. Further, since the trades are done sequentially, the trader is allowed to observe all past trades in the market and update his private information based on this information.

Traders are allowed to report their beliefs on the outcome of the event by allowing them to buy and sell securities whose value depends on the outcome of this future event. This will effect the state of the market, thus updating the predictive probabilities for the event. For example, the Iowa Electronic Market predicts the outcome of a presidential election by allowing one to trade on securities that pay off only if the Democratic candidate wins, or only if the Republican candidate wins. Traders can see the past history of trades, so the price at which a current trader is willing to buy and sell these securities can be interpreted as an aggregate ``market probability'' of a particular candidate winning the election.

Currently, most deployed prediction markets have binary or finite discrete outcome spaces. This means that the outcome of every event can be characterized by a finite set of mutually exclusive exhaustive outcomes. The securities in these market are usually tied to each outcome so that buying (respectively, selling) a security related to an outcome, indicates a belief in an increased (respectively, decreased) probability of occurrence of the outcome.

One form of prediction markets, which is rapidly gaining in popularity, is the market scoring rule~\cite{hanson03}. A market scoring rule considers all trades as a single chronological sequence. Traders earn rewards proportional to the incremental reduction in prediction loss caused by their trades in comparison to the previous trade. In other words, their rewards depend on the change in market probabilities caused by their trade, as well as on the eventual outcome. Thus, each trader has an 
incentive to minimize the prediction loss. In this format, the {\em market maker} who runs the market can suffer an overall loss, but Hanson~\cite{hanson03} showed that, for market scoring rules on finite outcome spaces, the loss of the market maker can be bounded.

%Currently, prediction markets are predominantly designed for predicting finite outcomes. In \cite{Gao} the authors extend the finite outcome prediction market to an infinite outcome space. For the logarithmic market scoring rule (LMSR) they show that the market maker's worst case loss in unbounded for an infinite outcome space.

Directly applying techniques used for binary outcome spaces to large outcome spaces is computationally infeasible. In fact, it has been shown that computing and updating prices for combinatorial outcomes spaces is $\#$P-hard \cite{Chen08}. Infinite outcome spaces pose a particular challenge in prediction markets since it is not clear how to define securities in such spaces. The most direct way of discretizing the space may lead to loss of information from the traders. Extending prediction market techniques to this space has been a subject of recent interest. Various techniques to bound market maker loss in both a combinatorial and infinite outcome space have been studied (see, for instance, \cite{Abernethy11,OthmanS11}).

In this paper, we 
use a powerful tool from statistics for handling infinite outcome spaces: the use of parametric distributions.
Specifically, we consider markets for possibly infinite outcome spaces, with the restriction that the true distribution over 
the outcomes belongs to an exponential family of distributions.

For such parametric families, we show that we can design a prediction market such that, when traders are strategically maximizing their own expected profit but are not adversarial, the market state can provably be interpreted as the maximum likelihood estimate of the natural parameters of the distribution. Also, this market has the desirable property of being {\em arbitrage-free}: There is no trade or sequence of trades that traders can exploit to make a guaranteed profit.

Next, we consider a more realistic situation in which their are both adversarial and strategic traders. We present a budget limited version of the prediction market that can be shown to suffer limited damage under concerted attacks by adversarial traders. We also exploit an information-theoretic interpretation of the particular form of the scoring rule that we use to show that an informative trader will eventually be unconstrained and he will be allowed to carry out an unrestricted trade. Taken together, we will have shown that while the market suffers limited damage from malicious traders, it is also able to make use of all the information from informative traders in the long run.

Apart from its value as a new form of prediction market, we believe that this connection between prediction markets and exponential families can be exploited further to design high performance online learning algorithms. In section~\ref{sec:conclusion}, we suggest future work in this direction. %Preliminary results were reported in \cite{KuttySami11}.

\subsection{Related Work}
Designing prediction markets to handle a large outcome space is an active area of research.  In \cite{Chen08}, the authors use a restricted betting language to design efficient markets for a combinatorial outcome space. This technique is generalized by \cite{Pennock11}. \cite{Gao09} consider extending various automated market makers to an infinite outcome space. For the logarithmic market scoring rule they show that unbounded market maker loss can result in this setting.   \cite{Abernethy11} and \cite{OthmanS11} specify frameworks under which they design cost function based markets that satisfy the desirable property of bounded market maker loss even in infinite outcome spaces.

The connection between machine learning and prediction markets has been studied previously. \cite{Chen10} and \cite{Abernethy11} have previously explored the connection to learning algorithms to inform the design and understanding of prediction markets. In particular, \cite{Chen10}  consider the correspondence between prediction markets with market scoring rules and the Follow the Regularized Leader algorithm proposed by \cite{Kalai05} and thus provide insight into the aggregation mechanism of a prediction market.

Independently of this work, Beygelzimer et al.~\cite{BLP12} have shown that, for a particular form of binary prediction markets and traders with log-utility, the long-run dynamics of trading activity and budgets over many prediction markets lead the markets to satisfy a bounded regret property with respect to the best single trader. Our results here, and the future work we have suggested, form a program to prove bounded regret properties of budget-limited prediction markets under much more general conditions.

\section{EXPONENTIAL FAMILIES}
A number of popular distributions are members of a class of a distributions called the exponential family. For instance, exponential, beta, Bernoulli and Poisson distributions are all members of the exponential family of distributions. %In fact, these are specific examples of exponential family distributions that have finite support. We restrict ourselves to these kinds of distributions here.
%exponential, log-normal, gamma, chi-squared, beta, Dirichlet, Bernoulli, categorical, Poisson, geometric, inverse Gaussian, von Mises
In this work, we are concerned with outcomes that are drawn from an exponential family distribution.

A member of an exponential family $\family$, $P_{\betavec} \in \family$ assigns a probability density (with respect to some base measure) over $x$ defined over a space $\mathcal{X}$ and has the following form: 
\[
  P_{\betavec}(x) = e^{\betavec . \suff(x) -\lpf(\betavec)}
\]
Note that $\mathcal{X}$ could be a multi-dimensional space.

Here, the set of natural parameters $\betavec$ is drawn from some space $\mathcal{B}$. $\lpf(\betavec)$ is known as the {\em log-partition function} and is defined as 
 $$\lpf(\betavec)\defeq\log\int e^{\betavec . \suff(x)}\mbox{d}x$$

The vector of sufficient statistics $\suff(x)$ is some function of the data and is defined differently for different members of the exponential family.

For example, the Bernoulli distribution is a discrete probability distribution that is a member of the exponential family. It takes value 1 with  probability $p$ and value 0 with  probability $1-p$. In this case, the vector of sufficient statistics,  natural parameters and log-partition function is given by:
\begin{eqnarray*}
\suff(x)&=&x\\
\betavec&=&\log\frac{p}{1-p}\\
\lpf(\betavec)&=&-\log(1-p)
\end{eqnarray*}

%Let us consider the univariate Gaussian distribution as a particular example. The univariate Gaussian is defined as 
%$$\mathcal{N}_{\mu,\sigma^2}(x)= \frac{1}{\sqrt{2\pi \sigma^2}}\exp\left\{-\frac{(x-\mu)^2}{2\sigma^2}\right\}$$
%
%In this case, the vector of sufficient statistics,  natural parameters and log-partition function is 
%\begin{eqnarray*}
%\suff(x)&=&\begin{pmatrix} x\\ x^2\end{pmatrix}\\
%\betavec&=&\begin{pmatrix} \beta_1\\ \beta_2\end{pmatrix}=\begin{pmatrix} \frac{\mu}{\sigma^2}\\ \frac{-1}{2\sigma^2}\end{pmatrix}\\
%\lpf(\betavec)&=& \sqrt{\frac{-\beta_2}{\pi}} \exp\left(\frac{\beta_1^2}{4\beta_2}\right)
%\end{eqnarray*}

\section{PREDICTION MARKETS}
A prediction market is a market where traders bet on the outcome of an event and are paid according to the accuracy of their forecast. The current market state indicates the net prediction on the probability of an event. Let $i\in\{1,\ldots,N\}$ be the possible (mutually exclusive, exhaustive) outcomes of the event. Let $\pvec$ indicate the current market state. Traders in the market influence the value of $\pvec$ and hence the predictive probability of the event.

Market scoring rules were developed as a way to allow sequential participation in a prediction market. In the Log Market Scoring Rule (LMSR), a trader who moves the current market estimate of an actual outcome $i$ from $p_{i-1}$ to $p_i$ receives a (possibly negative) payoff of $\log\frac{p_i}{p_{i-1}}$.

In a cost-function based prediction market, a trader pays for the movement of the market probability upfront \cite{hanson03}. This movement is associated with the purchase of securities. The cost of purchase is determined by a cost function. The trader is then paid according to a payoff function. %Both of these functions are non-negative.
Ideally, the cost function is such that it covers any possible loss that the traders may incur on the trade, thereby preventing any bankruptcy problems at the time of payoff settlement.
